Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-27 paper.tex
Update paper.tex
@jeromekelleher :
 the line numbers refer to the github version of paper.tex 

line 217: added  ',and because the mathematical construction separates the mutation process from the ancestral process.'

line 873: 'human-like' sounds a bit funny 
how about 'For values of parameters (recombination rate, mutation rate)  that have been estimated for humans ...' 

line 1452: 'Community-based approach' sounds more appropriate 

these are only  suggestions, either way is fine

The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is an efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from an idealised population model, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and used these simulations to characterise an analytically
intractable distribution.
This inherent efficiency, and the great utility of simulations for a wide
range of purposes, has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

@@

In human-like parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the population scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithm tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
As mentioned above, common ancestor events in which the ancestral material
of two lineages is merged only result in coalescences in the marginal trees
if their ancestral segments overlap. If there is no overlap, the merged
segments represent an ancestral chromosome that is a genetic ancestor
of the two lineages, but not the most recent common genetic ancestor
at any location along the genome.
When this happens, the merged lineage carries ``trapped'' genetic material
that is not ancestral to any samples, but where recombinations can still
occur~\citep{wiuf1999recombination}.
For large $\rho$, recombination events in trapped
ancestral material will dominate, and so we can use this as a proxy for the
overall number of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
for $n$ samples.
As discussed in the \nameref{app-time-complexity} appendix, the quadratic
dependence of simulation running time on $\rho$ implied
by \eqref{eqn-hsw-bound} is well supported
by observations, and provides a useful means of predicting how long
a particular simulation might require.

@@

\subsection*{Integration with forward simulators}
A unique feature of \msprime\ is its ability to simulate
genetic ancestries by extending an existing partial
genetic ancestry. Given a tree sequence that
is complete up until time $t$ ago as input
(where marginal trees may or may not have fully coalesced),
\msprime\ can efficiently obtain the segments of ancestral material
present at this time, and then run the simulation backwards in time from there.
This allows a simulated ancestry to be produced by any
number of different processes across disjoint time slices.
In practice this feature is used to ``complete''
forwards-time ancestry simulations~\citep{kelleher2018efficient}
that may have not fully coalesced. This process
(``recapitation'') can be orders of magnitude faster than
the standard approach of neutral burn-in; see
\cite{haller2018tree} for more details and examples.
This interoperability between simulators, where a partial ancestry
simulation produced by \SLiM~\citep{haller2019slim}
or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
by another simulator, with complete information
retained---at scale---is unprecedented.
There may be an opportunity for
other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
to leverage the tree sequence data format and associated tools.

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
These tests are run
automatically on different operating systems on each pull request
(where a contributor proposes a code change), using standard Continuous
Integration (CI) methodology. Other CI services
check for common errors, code formatting issues, and produce reports on
the level of test coverage for the proposed change.

>>>>

The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is an efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from an idealised population model, without explicitly representing
the population in memory or stepping through the generations, and because the mathematical construction separates the mutation process from the ancestral process.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and used these simulations to characterise an analytically
intractable distribution.
This inherent efficiency, and the great utility of simulations for a wide
range of purposes, has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

@@

%%'human-like' sounds a bit funny
In human-like  parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the population scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithm tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
As mentioned above, common ancestor events in which the ancestral material
of two lineages is merged only result in coalescences in the marginal trees
if their ancestral segments overlap. If there is no overlap, the merged
segments represent an ancestral chromosome that is a genetic ancestor
of the two lineages, but not the most recent common genetic ancestor
at any location along the genome.
When this happens, the merged lineage carries ``trapped'' genetic material
that is not ancestral to any samples, but where recombinations can still
occur~\citep{wiuf1999recombination}.
For large $\rho$, recombination events in trapped
ancestral material will dominate, and so we can use this as a proxy for the
overall number of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
for $n$ samples.
As discussed in the \nameref{app-time-complexity} appendix, the quadratic
dependence of simulation running time on $\rho$ implied
by \eqref{eqn-hsw-bound} is well supported
by observations, and provides a useful means of predicting how long
a particular simulation might require.

@@

\subsection*{Integration with forward simulators}
A unique feature of \msprime\ is its ability to simulate
genetic ancestries by extending an existing partial
genetic ancestry. Given a tree sequence that
is complete up until time $t$ ago as input
(where marginal trees may or may not have fully coalesced),
\msprime\ can efficiently obtain the segments of ancestral material
present at this time, and then run the simulation backwards in time from there.
This allows a simulated ancestry to be produced by any
number of different processes across disjoint time slices.
In practice this feature is used to ``complete''
forwards-time ancestry simulations~\citep{kelleher2018efficient}
that may have not fully coalesced. This process
(``recapitation'') can be orders of magnitude faster than
the standard approach of neutral burn-in; see
\cite{haller2018tree} for more details and examples.
This interoperability between simulators, where a partial ancestry
simulation produced by \SLiM~\citep{haller2019slim}
or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
by another simulator, with complete information
retained---at scale---is unprecedented.
There may be an opportunity for
other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
to leverage the tree sequence data format and associated tools.
%% maybe 'Community-based approach'  
\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality, correctness, good documentation, and inclusive development.
As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
These tests are run
automatically on different operating systems on each pull request
(where a contributor proposes a code change), using standard Continuous
Integration (CI) methodology. Other CI services
check for common errors, code formatting issues, and produce reports on
the level of test coverage for the proposed change.
